### 2021-11-16 documents/sections/verbs.tex
'to go' and 'to come'

\subsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb is reconstructed by \textcite{gildea2007greenberg} as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
It is true that many reflexes are clearly \obj{t}-initial, for example \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}.
However, once one considers all forms of the various reflexes of this verb \pcref{tab:go}, an initial vowel \rc{ɨ} must clearly be reconstructed -- in contrast to unambiguously C-initial \rc{ka[ti]} \qu{to say}.%
\footnote{As indicated by the brackets in \cref{tab:go}, there are many languages where the initial vowel is only present in some forms.
Also, the prefix-verb boundary in many inflected forms like e.g.\ \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since an epenthetic \obj{ɨ} breaks up potential CC clusters.}
This verb was not affected by the any of the extensions discussed in \cref{sec:extensions}.

@@

This verb is reconstructed as \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a somewhat more complex story.
Crucially, the \rc{ət} part of their reconstruction is not reflected in the majority of forms, and many languages ostensibly have reflexes of \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ} for the \rc{epɨ} part.
We analyze these forms as going back to a \PC verb of the form \rc{(ət-)jəpɨ}, morphologically segmentable into a detransitivizer and a root \rc{jəpɨ}.

Evidence for the originally \rc{j}-initial nature of the root is found in the Pemongan languages and \kaxui, although the simultaneous presence of \obj{oohɨ} in \kaxui raises the question of whether the \obj{j} is part of the root in \obj{johɨ}.
This form only occurs in the Progressive \exref{kaxprog.kax-82}, and indeed \obj{j(-)} may be a reflex of \rc{i-w-} \qu{\gl{3}-\gl{s_a_}-}, as the \gl{s_a_} class marker \rc{w-} is present with other person values \exref{kaxprog.kax-54}.
However, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial \gl{s_a_} verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, we are confident that this \obj{j} is indeed part of the root, rather than an outcome of \rc{i-w-}, which in turn allows us to reconstruct \rc{j} back to \PC.

\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
\a<kax-82>
\begingl
\gla johɨ-rɨ//
\glb \gl{3}.come-\gl{prog}//
\glft \qu{S/he is coming.}//
\endgl
\a<kax-54>
\begingl
\gla o-w-ohɨ-rɨ//
\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
\glft \qu{You are coming.}//
\endgl
\a<kax-77>
\begingl
\gla i-nkɨ-rɨ//
\glb \gl{3}-sleep-\gl{prog}//
\glft \qu{S/he is sleeping.}//
\endgl
\a<kax-83>
\begingl
\gla Ø-osone-rɨ//
\glb \gl{3}-dream-\gl{prog}//
\glft \qu{S/he is dreaming.}//
\endgl
\xe
%
Most forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} do not show evidence for this \rc{j}.
However, we interpret the \obj{i} in the \akawaio form \obj{əsipɨ} as additional evidence of the sequence \rc{jə}, which has the same outcome \obj{i} in the \macushi reflex of bare \rc{jəpɨ}, \obj{ipɨ}.

Turning to the many forms apparently reflecting \rc{əpɨ} and \rc{epɨ}, we find that both are distributed widely in the family, sometimes even co-occurring in the same language.
A unifying account of these forms requires the root \rc{jəpɨ} to undergo two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
Both phenomena are found in other contexts throughout the family \parencite{meira2010origin}.
For this particular verb, these sound changes appear to have applied irregularly, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as reflecting an earlier diphthong \rc{əe}.

@@

Our interpretation of the \rc{ət} portion as a detransitivizer is primarily based on its form, as well as its paradigmatically conditioned occurrence in some languages.
Although the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
Also, \waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \panare \exref{pan-128}, \arara, and \trio.

@@

While these sound changes and the addition of \rc{ət-} do account for the majority of the forms in \cref{tab:come},\footnote{
Apart from aforementioned \akuriyo \obj{eepɨ}, another irregular form is \apalai \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}.
While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-} in \apalai.
The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}.
Alternatively, \apalai \obj{oepɨ} could be a fossilized loan from \wayana, which has replaced its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Apart from the distribution of \rc{əpɨ} and \rc{epɨ}, forms with and without \rc{ət-} can be found within the same language, usually conditioned by different prefixes.
This was briefly discussed in \cref{sec:pekodian} for \arara and in \cref{sec:taranoan} for \trio (and \PTir and \PTar).
To illustrate, the \trio \setone paradigm shows a reflex of \rc{ətepɨ} (< \rc{ətjəpɨ}) for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}%
\footnote{It should be noted that the paradigmatic distributions of forms with and without \rc{ət-} in different languages share no discernible patterns.}

@@

As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Both reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ} and/or \rc{epɨ} (\akuriyo) resisted the introduction of \rc{t-} in \PTir.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come} and was thus not a potential target of innovative \rc{k-}.

>>>>

\subsection{\rc{ɨtə[mə]} \qu{to go}}
\label{sec:go}
This verb is reconstructed by \textcite{gildea2007greenberg} as \rc{tə[mə]}, like \rc{ka[ti]} \qu{to say} with a fleeting second syllable.
It is true that many reflexes are clearly \obj{t}-initial, for example \hixka \obj{ntoje} \qu{he went} \parencite[27]{hixkaryanaderby1985}, \trio \obj{təkə} \qu{go!} \parencite[246]{triomeira1999}, or \wayana \obj{kuptəm} \qu{we went} \parencite[195]{wayanatavares2005}.
However, once one considers all forms of the various reflexes of this verb \pcref{tab:go}, an initial vowel \rc{ɨ} must clearly be reconstructed -- in contrast to unambiguously C-initial \rc{ka[ti]} \qu{to say}.%
\footnote{As indicated by the brackets in \cref{tab:go}, there are many languages where the initial vowel is only present in some forms.
Also, the prefix-verb boundary in many inflected forms like e.g.\ \trio \obj{wɨtənne} or \arara \obj{wɨdolɨ} \qu{I went} \parencites[43]{triomeira1999}[153]{alves2017arara} is ambiguous, since an epenthetic \obj{ɨ} breaks up potential CC clusters.
Still, when one considers unambiguous forms, the contrast with \rc{ka[ti]} becomes very clear.}
This verb was not affected by the any of the extensions discussed in \cref{sec:extensions}.

@@

This verb is reconstructed as \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a somewhat more complex story.
Crucially, the \rc{ət} part of their reconstruction is not reflected in the majority of forms, and many languages ostensibly have reflexes of \rc{əpɨ}, \rc{jepɨ}, or \rc{jəpɨ} for the \rc{epɨ} part.
We analyze these forms as going back to a \PC verb of the form \rc{(ət-)jəpɨ}, morphologically segmentable into a detransitivizing prefix and a root \rc{jəpɨ}.

Evidence for the originally \rc{j}-initial nature of the root is found in the Pemongan languages and \kaxui, although the coexistence of \obj{oohɨ} and \obj{johɨ} in \kaxui raises the question of whether the \obj{j} in the latter form is actually part of the root.
The form in question only occurs in the Progressive \exref{kaxprog.kax-82}, and it seems possible that \obj{j(-)} is actually a reflex of \rc{i-w-} \qu{\gl{3}-\gl{s_a_}-}, since the \gl{s_a_} class marker \rc{w-} is present with other person values \exref{kaxprog.kax-54}.
However, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial \gl{s_a_} verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, the best explanation for this \obj{j} is that is indeed part of the root, rather than an outcome of \rc{i-w-}, which in turn allows us to safely reconstruct \rc{j} back to \PC.

\pex<kaxprog>\kaxui \perscomm{Spike Gildea}
\a<kax-82>
\begingl
\gla johɨ-rɨ//
\glb \gl{3}.come-\gl{prog}//
\glft \qu{S/he is coming.}//
\endgl
\a<kax-54>
\begingl
\gla o-w-ohɨ-rɨ//
\glb \gl{2}-\gl{s_a_}-come-\gl{prog}//
\glft \qu{You are coming.}//
\endgl
\a<kax-77>
\begingl
\gla i-nkɨ-rɨ//
\glb \gl{3}-sleep-\gl{prog}//
\glft \qu{S/he is sleeping.}//
\endgl
\a<kax-83>
\begingl
\gla Ø-osone-rɨ//
\glb \gl{3}-dream-\gl{prog}//
\glft \qu{S/he is dreaming.}//
\endgl
\xe
%
Most of the longer forms corresponding to \posscite{gildea2007greenberg} \rc{ətepɨ} do not show evidence for this \rc{j}, but we identify the \obj{i} in the \akawaio form \obj{əsipɨ} as a reflex of the sequence \rc{jə}, which has the same outcome in the \macushi reflex of bare \rc{jəpɨ}, \obj{ipɨ}.

Turning to the many forms seemingly reflecting \rc{əpɨ} and \rc{epɨ}, we find that both are distributed widely in the family, sometimes even co-occurring in the same language.
A unifying account of these forms requires the root \rc{jəpɨ} to undergo two major sound changes: \begin{inlinelist}
 \item \rc{j}-loss
 \item \rc{ə}-umlaut after \rc{j}
 \end{inlinelist}.
Both phenomena are found in other contexts throughout the family \parencite{meira2010origin}.
For this particular verb, these sound changes appear to have applied irregularly, and not always in the same order.
For example, the \kalina form \obj{opɨ} can only be explained if \rc{j} was lost before the umlaut of \rc{ə} to \rc{e}, which would have been triggered by \rc{j}.
On the other hand, forms like \maqui \obj{ehə} must be the result of \phonl{\rc{ə}}{\rc{e}}{\rc{j}}, with subsequent loss of \rc{j}.
The \akuriyo form \obj{eepɨ} looks like a reflex thereof as well, but the length is unexpected, and is analyzed by \textcite[]{meira1998proto} as reflecting an earlier diphthong \rc{əe}.

@@

Our interpretation of the \rc{ət} portion as a detransitivizer is primarily based on its form, as well as its paradigmatically conditioned occurrence in some languages, discussed shortly.
Although the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, the \PC \gl{s_p_} verb \rc{wɨnɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
Also, \waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \arara, \trio, and \panare \exref{pan-128}.

@@

While a root \rc{jəpɨ}, aforementioned sound changes, and the addition of \rc{ət-} do account for the majority of the forms in \cref{tab:come},\footnote{
Apart from aforementioned \akuriyo \obj{eepɨ}, another diachronically irregular form is \apalai \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}.
While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-} in \apalai.
The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}.
Alternatively, \apalai \obj{oepɨ} could be a fossilized loan from \wayana, which has replaced its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of the intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Apart from the distribution of \rc{əpɨ} and \rc{epɨ}, forms with and without \rc{ət-} can be found within the same language, usually conditioned by different prefixes.
This was briefly discussed in \cref{sec:pekodian} for \arara (and \PPek) and in \cref{sec:taranoan} for \trio (\PTar).
To illustrate, the \trio \setone paradigm shows a reflex of \rc{ətepɨ} (< \rc{ətjəpɨ}) for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}%
\footnote{It should be noted that the paradigmatic distributions of forms with and without \rc{ət-} in different languages share no discernible patterns.}

@@

As discussed in \cref{sec:pekodian}, innovative \rc{k-} was introduced on the \ikpeng and \bakairi reflexes of \rc{ətjəpɨ}, but not on the \arara reflex of \rc{jəpɨ}.
Both reflexes of \rc{ətjəpɨ} (\trio) and of \rc{ətjəpɨ} and/or \rc{jəpɨ} (\akuriyo) resisted the introduction of \rc{t-} in \PTir.
\carijo \obj{ehɨ} shows innovative \obj{j-}, rather than conservative \obj{w-} \exref{car-18}.
It is unknown whether there is a \yukpa reflex of this verb, and it was fully replaced in \PWai by \rc{omokɨ} \qu{to come} and was thus not a potential target of innovative \rc{k-}.
